sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields

### 2023-01-25 doc/paper-jss/sdmTMB-cover.tex
Edits

Functionality: This package represents a spatial analog of the \pkg{glmmTMB} \proglang{R} package, with additional features beyond random fields (penalized smoothers as in \pkg{mgcv}, priors as in \pkg{INLA}, threshold/breakpoint models), spatial anisotropy (directionally dependent spatial correlation), and methods for model comparison (cross-validation, stacking).

Accessibility: \pkg{sdmTMB} provides a much more usable interface than alternatives, particular for users who may be familiar with the syntax found in \proglang{R}'s \texttt{glm()}, \pkg{glmmTMB}, \pkg{lme4}, or \pkg{mgcv} packages. Existing packages, such as \pkg{INLA} or \pkg{inlabru} are fundamentally Bayesian and require users to deal with the complexity of priors. While \pkg{sdmTMB} can incorporate priors and extend to full Bayesian inference via \pkg{Stan}, by default estimation is done using the simplicity and speed of maximum likelihood.

Speed: The package combines the SPDE approximation with estimation in the \pkg{TMB} \proglang{R} package for computational efficiency. Across our comparisons, \pkg{sdmTMB} was 6--24 times faster than the \pkg{INLA} or \pkg{inlabru} packages, the next fastest related software.

>>>>

Functionality: This package represents a spatial analog of the \pkg{glmmTMB} \proglang{R} package, with additional features beyond standard random fields (penalized smoothers as in \pkg{mgcv}, priors as in \pkg{INLA}, threshold/breakpoint models), spatial anisotropy (directionally dependent spatial correlation), and methods for model comparison (cross-validation, stacking).

Accessibility: \pkg{sdmTMB} provides a much more usable interface than alternatives, particularly for users who may be familiar with the syntax found in \proglang{R}'s \texttt{glm()}, \pkg{glmmTMB}, \pkg{lme4}, or \pkg{mgcv} packages. Existing packages, such as \pkg{INLA} or \pkg{inlabru} are fundamentally Bayesian and require users to deal with the complexity of priors. While \pkg{sdmTMB} can incorporate priors and extend to full Bayesian inference via \pkg{Stan}, by default estimation is done using the simplicity and speed of maximum likelihood.

Speed: The package combines the SPDE approximation with estimation in the \pkg{TMB} \proglang{R} package for computational efficiency. Across our comparisons, \pkg{sdmTMB} was 6--24 times faster than the \pkg{INLA} or \pkg{inlabru} packages, the software with the most similar capabilities for this class of models.
